---
title: "Introduction to rankedBeliefR"
author: "rankedBeliefR Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to rankedBeliefR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Don't evaluate code during package build
)
```

## Overview

The `rankedBeliefR` package provides R bindings for the ranked_belief C++ library, which implements lazy ranking functions for epistemic reasoning. This package allows you to work with ranked beliefs—probability distributions where outcomes are ordered by plausibility ranks rather than precise probabilities.

**Note:** This package is a proof-of-concept demonstrating the C API of the ranked_belief library. It is not intended for production use but serves as a validation of the C API design.

## Key Concepts

- **Ranking**: A collection of values, each with a non-negative rank indicating exceptionality (0 = normal, higher = more exceptional)
- **Lazy evaluation**: Rankings are computed on-demand, enabling infinite structures
- **Monadic composition**: Use `rb_merge_apply_int` to compose dependent rankings

## Basic Operations

### Creating Rankings

```{r create_rankings}
library(rankedBeliefR)

# Singleton ranking (one value at rank 0)
single <- rb_singleton_int(42L)
rb_take_n_int(single, 5)
rb_free(single)

# From arrays (values with different ranks)
ranking <- rb_from_array_int(
  values = c(1L, 2L, 3L),
  ranks = c(0, 1, 2)
)
rb_take_n_int(ranking, 5)
rb_free(ranking)
```

### Transforming Rankings

```{r transform}
# Map: apply function to each value
original <- rb_from_array_int(values = c(1L, 2L, 3L), ranks = c(0, 1, 2))
doubled <- rb_map_int(original, function(x) x * 2L)
rb_take_n_int(doubled, 5)
rb_free(original)
rb_free(doubled)

# Filter: keep only values matching predicate
numbers <- rb_from_array_int(values = 1L:10L, ranks = rep(0, 10))
evens <- rb_filter_int(numbers, function(x) x %% 2 == 0)
rb_take_n_int(evens, 10)
rb_free(numbers)
rb_free(evens)
```

### Combining Rankings

```{r combine}
# Merge: interleave two rankings by rank
r1 <- rb_from_array_int(values = c(1L, 3L), ranks = c(0, 2))
r2 <- rb_from_array_int(values = c(2L, 4L), ranks = c(1, 3))
merged <- rb_merge_int(r1, r2)
rb_take_n_int(merged, 5)
rb_free(r1)
rb_free(r2)
rb_free(merged)
```

## Dependent Rankings Example: Beer and Peanuts

This example shows how to model dependent variables where peanut consumption depends on beer consumption.

```{r beer_peanuts}
# Helper: normally FALSE, exceptionally TRUE
nrm_exc_bool <- function() {
  rb_from_array_int(values = c(0L, 1L), ranks = c(0, 1))
}

# Peanuts depend on beer
peanuts_given_beer <- function(beer_val) {
  if (beer_val == 1L) {
    # If drinking beer, normally eat peanuts (1), exceptionally don't (0)
    rb_from_array_int(values = c(1L, 0L), ranks = c(0, 1))
  } else {
    # If not drinking beer, don't eat peanuts
    rb_singleton_int(0L)
  }
}

# Encode (beer, peanuts) pair as single integer
encode_pair <- function(beer, peanuts) as.integer(beer * 2 + peanuts)
decode_pair <- function(val) c(beer = val %/% 2, peanuts = val %% 2)

# Create dependent ranking using rb_merge_apply_int (monadic bind)
beer_dist <- nrm_exc_bool()

combined <- rb_merge_apply_int(beer_dist, function(beer_val) {
  peanuts_dist <- peanuts_given_beer(beer_val)
  
  # Map to encoded pairs
  rb_merge_apply_int(peanuts_dist, function(peanuts_val) {
    rb_singleton_int(encode_pair(beer_val, peanuts_val))
  })
})

# Sample results
samples <- rb_take_n_int(combined, 10)
cat("Possible scenarios (beer=0/1, peanuts=0/1):\n")
for (i in seq_len(nrow(samples))) {
  pair <- decode_pair(samples$value[i])
  cat(sprintf("  beer=%d, peanuts=%d (rank %g)\n", 
              pair['beer'], pair['peanuts'], samples$rank[i]))
}

rb_free(beer_dist)
rb_free(combined)
```

**Interpretation:**
- Rank 0: No beer, no peanuts (most normal)
- Rank 1: Beer and peanuts (one exceptional event: drinking beer)
- Rank 2: Beer but no peanuts (two exceptional events)

## Memory Management

Always call `rb_free()` on ranking objects when done to prevent memory leaks:

```{r memory, eval=FALSE}
ranking <- rb_singleton_int(1L)
# ... use ranking ...
rb_free(ranking)  # Clean up
```

## Next Steps

- See `vignette("boolean-circuit")` for a practical diagnosis example
- See `vignette("recursive-rankings")` for infinite lazy structures
- Check `inst/examples/` for more code samples

## API Reference

Core functions:
- `rb_singleton_int(value)`: Create singleton ranking
- `rb_from_array_int(values, ranks)`: Create ranking from arrays
- `rb_map_int(ranking, fn)`: Transform values
- `rb_filter_int(ranking, predicate)`: Filter values
- `rb_merge_int(r1, r2)`: Combine rankings
- `rb_merge_apply_int(ranking, fn)`: Monadic bind (composition)
- `rb_observe_value_int(ranking, value)`: Condition on observation
- `rb_take_n_int(ranking, n)`: Sample first n elements
- `rb_first_int(ranking)`: Get first element
- `rb_is_empty(ranking)`: Check if empty
- `rb_free(ranking)`: Free memory
